Well-quasi-orderings on word languages

### 2024-09-19 src/introduction.tex
work on intro/contribution

\AP
Note that in the case of finite words and finite trees, the precise choice of
ordering on words and trees is crucial to ensure that the resulting structure
is \kl{well-quasi-ordered}. The celebrated result of Higman states that the set
of finite words over an ordered alphabet $(X, \preceq)$ is
\kl{well-quasi-ordered} by the so-called \kl{subword embedding relation}
\cite{HIG52}. Let us recall that the \kl{subword relation} for words over $(X,
\preceq)$ is defined as follows: a word $u$ is a \intro{subword} of a word $v$,
written $u \intro*\higleq v$,
if there exists an increasing function $f \colon \{1, \ldots, |u|\} \to \{1,
\ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all $i \in \{1, \ldots,
|u|\}$.

\begin{figure}
    \centering
    \begin{tabular}{p{3cm}p{8cm}}
    Prefix Relation
    &
    \includestandalone[width=\linewidth]{fig/prefix-embedding-standalone}
    \\
    Infix relation
    &
    \includestandalone[width=\linewidth]{fig/infix-embedding-standalone}
    \\
    Subword relation
    &
    \includestandalone[width=\linewidth]{fig/subword-embedding-standalone}
    \end{tabular}
    \caption{A simple representation of the \kl{subword relation},
        \kl{prefix relation},
        and \kl{infix relation},
        on the alphabet $\set{a,b}$ for words of
        length at most $3$. The figures are Hasse Diagrams,
        representing the successor relation of the order.
        Furthermore, we highlight in red relations that are added
        when moving from the prefix relation to the infix one,
        and to the infix relation to the subword one.}
    \label{word-embeddings:fig}
\end{figure}

\AP However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. Let us restrict
ourselves to a simplified case where the alphabet is finite and equipped with
the equality relation. In this setting, the three alternatives we consider are
the \intro{prefix relation} --- $u \intro*\prefleq v$ if there exists $w$ with
$uw = v$ ---, the \intro{suffix relation} --- $u \intro*\suffleq v$ if there
exists $w$ such that $wu = v$ ---, and the \intro{infix relation} --- $u
\intro*\infleq v$ if there exists $w_1,w_2$ such that $w_1 u w_2 = v$ ---. Note
that these three relations straightforwardly generalize to infinite
quasi-ordered alphabets. Unfortunately, it is an easy check that all of these
constructions are not well-quasi-ordered as soon as the alphabet contains two
non-equivalent letters. This is not the end of the road, because some
\emph{subsets} of $\Sigma^*$ might still be \kl{well-quasi-ordered} by these
relations, the simplest example being the case of finite sets of (finite) words
which are all \kl{well-quasi-ordered} regardless of the ordering considered.
This raises the question of characterizing the subsets $L \subseteq \Sigma^*$
that are \kl{well-quasi-ordered} with respect to the \kl{prefix relation}
(respectively, the \kl{suffix relation} or the \kl{infix relation}), and
providing suitable decision procedures.

\subparagraph{Contribution} \AP In this paper, we focus on words over a finite
alphabet $\Sigma$, and characterize subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation},
and the \kl{infix relation}. Let us remark that these word orderings are all
\intro{well-founded}, i.e., they do not contain infinite \intro{decreasing
sequences} $\seqof{x_n}$ such that $x_{n+1} \prec x_n$ for all $n \in \Nat$.
Let us recall that the key difference between \kl{well-quasi-orderings} and the
maybe more usual notion of well-founded orderings is that the latter allows the
existence of infinite \intro{antichains}, that is, sets of pairwise
incomparable elements. In fact, a set $X$ is \kl{well-quasi-ordered} if and
only if it is \kl{well-founded} and does not contain infinite \kl{antichains}
\cite{KRU72}.  As a consequence, our study focuses on detecting infinite
\kl{antichains} in languages $L \subseteq \Sigma^*$. Remark that for the
relations we consider (including the \kl{subword relation}), the set of all
words of a fixed size $k$ is always an \kl{antichain}, and there are therefore
no bound on the maximal length of an antichain.

In the case of the \kl{prefix relation}, we provide a complete characterization
of languages $L \subseteq \Sigma^*$ that are \kl{well-quasi-ordered}, which are
the simplest possible languages: they are precisely the \emph{finite} unions of
\intro{chains}, that is, sets of words that are all comparable for the
\kl{prefix relation}. Note that \kl{chains} are the simplest possible
\kl{well-quasi-ordered} sets (they are totally ordered and well-founded) and it
is known that finite unions of \kl{well-quasi-ordered} sets are
\kl{well-quasi-ordered}. As a consequence, the above characterization states
that only the simplest possible \kl{well-quasi-ordered} sets are
\kl{well-quasi-ordered} by the \kl{prefix relation}. Let us furthermore
highlight that this characterization holds without any restriction on the
decidability of the language $L$ itself, but heavily relies on the assumption
that $\Sigma$ is finite. By symmetry, we obtain an analogue statement for the
\kl{suffix relation}.

In the case of the \kl{infix relation}, the situation is more complex, and we
show that \emph{any} countable quasi-ordering with finite initial segments can
be \kl{embedded} into the \kl{infix relation} of a language over a binary
alphabet $\Sigma \defined \set{a,b}$. As an example, on can identify
$(\Sigma^*, \higleq)$ with a subset of $(\Sigma^*, \infleq)$ where $\higleq$ is
the \kl{subword relation} and $\infleq$ is the \kl{infix relation}. This
strongly contrasts with the \kl{prefix relation}, and suggests that a
meaningful characterization of \kl{well-quasi-ordered} languages under the
\kl{infix relation} is unlikely to be found.

We circumvent this issue by adding some regularity assumptions on the languages
$L \subseteq \Sigma^*$ that we consider. This can be done by assuming that $L$
is recognized by a finite automaton, or more generally that $L$ is the language
of an \kl{amalgamation system}, a recently introduced notion that abstracts the
notion of systems whose runs can be suitably ``pumped'' \cite{ASZZ24}. As
examples \kl{amalgamation systems} covers the case of regular languages,
context free languages, or even languages recognized by ``vector addition
systems with state" \cite{ASZZ24}. Under this relatively mild assumption, we
also prove that \kl{well-quasi-ordered} languages under the \kl{infix relation}
are precisely the finite unions of \kl{chains}.

@@

\subparagraph{Outline} In \cref{prefixes:sec}, which is relatively
self-contained, we study the \kl{prefix relation} prove in \cref{prefixes:thm}
the characterization of \kl{well-quasi-ordered} languages by the \kl{prefix
relation}. In \cref{infixes-regular:sec}, we prove that any countable quasi-ordering
with finite initial segments can be embedded into the \kl{infix relation} of a
language (\cref{infix-embedding:thm}), we obtain the analogue of
\cref{prefixes:thm} for languages recognized by finite automata
(\cref{infix-finite-automata:thm}). Finally, in \cref{infixes-amalgamation:sec}, we
introduce \kl{amalgamation systems} and generalize
\cref{infix-finite-automata:thm} to this setting
(\cref{infix-amalgamation:thm}).

>>>>

\AP Note that in the case of finite words and finite trees, the precise choice
of ordering is crucial to ensure that the resulting structure is
\kl{well-quasi-ordered}. The celebrated result of Higman states that the set of
finite words over an ordered alphabet $(X, \preceq)$ is \kl{well-quasi-ordered}
by the so-called \kl{subword embedding relation} \cite{HIG52}. Let us recall
that the \kl{subword relation} for words over $(X, \preceq)$ is defined as
follows: a word $u$ is a \intro{subword} of a word $v$, written $u
\intro*\higleq v$, if there exists an increasing function $f \colon \{1,
\ldots, |u|\} \to \{1, \ldots, |v|\}$ such that $u_i \preceq v_{f(i)}$ for all
$i \in \{1, \ldots, |u|\}$.

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.48\textwidth}
    	\centering
    	\includestandalone[width=\linewidth]{fig/prefix-embedding-standalone}
    	\caption{Prefix Relation}
   	\end{subfigure}%
   	\hfill%
   	\begin{subfigure}[t]{0.48\textwidth}
   		\centering
   		\includestandalone[width=\linewidth]{fig/infix-embedding-standalone}
   		\caption{Infix Relation}
   	\end{subfigure}
   	\begin{subfigure}[t]{0.48\textwidth}
   		\centering
   		\includestandalone[width=\linewidth]{fig/subword-embedding-standalone}
   		\caption{Subword Relation}
   	\end{subfigure}

   	\caption{A simple representation of the \kl{subword relation},
        \kl{prefix relation},
        and \kl{infix relation},
        on the alphabet $\set{a,b}$ for words of
        length at most $3$. The figures are Hasse Diagrams,
        representing the successor relation of the order.
        Furthermore, we highlight in red relations that are added
        when moving from the prefix relation to the infix one,
        and to the infix relation to the subword one.}
    \label{word-embeddings:fig}
\end{figure}

\AP However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. Let us restrict
ourselves to a simplified case where the alphabet is finite and equipped with
the equality relation. In this setting, the three alternatives we consider are
the \intro{prefix relation} ($u \intro*\prefleq v$ if there exists $w$ with $uw
= v$), the \intro{suffix relation} ($u \intro*\suffleq v$ if there exists $w$
such that $wu = v$), and the \intro{infix relation} ($u \intro*\infleq v$ if
there exists $w_1,w_2$ such that $w_1 u w_2 = v$). Note that these three
relations straightforwardly generalize to infinite quasi-ordered alphabets.
Unfortunately, it is easy to see that all of these constructions are not
well-quasi-ordered as soon as the alphabet contains two non-equivalent letters:
for instance, the infinite sequence $a^n b^n a^n$ is \kl{well-quasi-ordered} by
the \kl{subword relation} but by neither the \kl{prefix relation}, nor the
\kl{suffix relation}, nor the \kl{infix relation}.

\AP While this dooms \kl[well-quasi-ordered]{well-quasi-orderedness} of these relations in the general case, there may be
\emph{subsets} of $\Sigma^*$ which are \kl{well-quasi-ordered} by these
relations. As a simple example, take the case of finite sets of (finite) words
which are all \kl{well-quasi-ordered} regardless of the ordering considered.
This raises the question of characterizing exactly which subsets $L \subseteq \Sigma^*$
are \kl{well-quasi-ordered} with respect to the \kl{prefix relation}
(respectively, the \kl{suffix relation} or the \kl{infix relation}), and
providing suitable decision procedures.

\subparagraph{Contribution} In this paper, we focus on words over a finite
alphabet $\Sigma$, and characterize subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation},
and the \kl{infix relation}. Specifically, for the \kl{prefix relation} and the \kl{suffix relation}, we are able to show the following decomposition into \intro{chains}, that is, totally ordered ascending sequences.

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\item A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{prefix relation} if and only if it is a finite union of \kl{chains} of the \kl{prefix relation}.
	\item A language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{suffix relation} if and only if it is a finite union of \kl{chains} of the \kl{suffix relation}.
\end{enumerate}
}

Note that \kl{chains} are the simplest possible
\kl{well-quasi-ordered} sets (they are totally ordered and well-founded) and it
is known that finite unions of \kl{well-quasi-ordered} sets are
\kl{well-quasi-ordered}. As a consequence, the above characterization states
that only the simplest possible \kl{well-quasi-ordered} sets are
\kl{well-quasi-ordered} by the \kl{prefix relation}. Let us furthermore
highlight that this characterization holds without any restriction on the
decidability of the language $L$ itself, but heavily relies on the assumption
that $\Sigma$ is finite.

\AP One way to measure the complexity of a \kl{well-quasi-ordered} set is to
consider its so-called \intro{ordinal invariants}. Several of these ordinal
measures have been developed for well-quasi-orders. The \intro{maximal order
type} (or m.o.t.) was originally defined by \cite{dejongh77} as the order type
of the maximal linearization of a well-quasi-ordered set. Then,
\cite{schmidt81} introduced the \intro{ordinal height} as the order type of a
maximal chain of a well-quasi-ordered set. Finally, \cite{kriz90b} crafted the
notion of \intro{ordinal width}, standardized the definitions of these three
\kl{ordinal invariants}, and proved numerical relationships between them. For
the \kl{prefix relation}, the \kl{suffix relation}, and the \kl{infix
relation}, the \kl{ordinal height} is always bounded by $\omega$, and the key
parameter to describe its complexity will be the \kl{ordinal width}, i.e.,
``how large the antichains are.''

\todo{rephrase}
\AP In the case of the \kl{prefix relation} and the \kl{suffix relation}, we
conclude from our characterization that the \kl{ordinal width} of a language
$L$ is either finite or $L$ is not \kl{well-quasi-ordered}. The straightforward
generalization of the results for the \kl{prefix relation} and the \kl{suffix
relation} to the \kl{infix relation} is not possible. For example, Kuske has
shown that any countable partial order with finite downward closures can be
embedded into it~\cite{DBLP:journals/ita/Kuske06}. In particular, this means
that the \kl{subword relation} can be embedded into the \kl{infix relation}.
The former being a \kl{well-quasi-ordered} with a very large \kl{ordinal width}
(see for example https://arxiv.org/pdf/2312.14587), this implies that such
"easy characterizations" will not be possible in general. However, not all is
lost: we show that with some mild restrictions we can recover much of the
simple structure of the \kl[prefix relation]{prefix} and \kl{suffix relation}:

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{2}
    \item A \emph{bounded} language $L \subseteq \Sigma^*$ is \kl{well-quasi-ordered} by the \kl{infix relation} if and only if it is a finite union of languages $S_i \cdot P_i$, where each $S_i$ is \kl{well-quasi-ordered} by the \kl{suffix relation} and each $P_i$ is \kl{well-quasi-ordered} by the \kl{prefix relation}.
\end{enumerate}
}

We illustrate how this result can be used to study languages recognized by
finite automata, context-free grammars, petri nets, and other systems. This is
possible by leveraging the recently introduced notion of \kl{amalgamation
systems} \cite{ASZZ24}.

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{2}
	\item A language $L \subseteq \Sigma^*$ recognized by an \kl{amalgamation system} is \kl{well-quasi-ordered} by the \kl{infix relation} if and only if it is a finite union of languages $S_i \cdot P_i$, where each $S_i$ is \kl{well-quasi-ordered} by the \kl{suffix relation} and each $P_i$ is \kl{well-quasi-ordered} by the \kl{prefix relation}.
\end{enumerate}
}

Finally, there has been recent interest in deciding whether a given order is a well-quasi-order. These questions have been investigated for example in the context of the verification of counter machines~\cite{DBLP:conf/fsttcs/FinkelG19} and in the context of logic~\cite{DBLP:journals/pacmpl/BergstrasserGLZ24}. We conclude our results with a contribution in this continuity.

{
\renewcommand{\labelenumi}{R\arabic{enumi}}
\begin{enumerate}
	\setcounter{enumi}{3}
	\item For languages of amalgamation systems, it is decidable whether they are \kl{well-quasi-ordered} by the \kl[prefix relation]{prefix}, \kl[suffix relation]{suffix}, or \kl{infix relation}.
\end{enumerate}
}

@@

\subparagraph{Outline} In
\cref{prefixes:sec}, which is relatively
self-contained, we study the \kl{prefix relation} and prove in
\cref{prefixes:thm} the characterization of \kl{well-quasi-ordered}
languages by the \kl{prefix relation}. In
\cref{infixes-regular:sec}, we
obtain the \kl[infix relation]{infix} analogue of \cref{prefixes:thm}
specifically for regular languages
(\cref{infix-finite-automata:thm}). Finally, we generalize this result to all
\kl{amalgamation systems} in \cref{infixes-amalgamation:sec}
in
(\cref{infix-amalgamation:thm}).
